### 2014-10-27 Paper1/paper1.tex

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Our actions often do not match our intentions when there are external disturbances. We derive a novel modeling approach for determining motor intent from a reaching motion that is disturbed by an unexpected force. We first show how to mathematically invert both feedforward (predictive) and feedback controls to obtain intent. We next examine the model’s sensitivity to parameter uncertainty, and test the approach using human reaching data. We found that sensitivity to parameter inaccuracies was on the order of square microns, and hence much lower than the inaccuracies seen in actual human motions. Further, the largest sensitivity arose mainly from uncertainty in our assumptions of joint stiffness. We also tested the hypothesis that this would predict straight-line intent in human targeted reaching movements, even when disturbed by various pulses of force. As expected, the intended trajectory showed no change from undisturbed reaching for 300 ms after the disturbance onset. Beyond this point, however, we detected a change in intent, even when the intention was near the target.  Knowing such an intent signal is broadly applicable in areas such as in enhanced human-machine interaction, the study of impaired intent in neural disorders, the real-time determination (and manipulation) of error in training, and in any complex system that embodies planning, such as brain machine interfaces, team sports, crowds, or swarms. In addition, observing intent as it changes might act as a window into the mechanisms of planning, correction, and learning.

@@

\section*{Introduction}
There are many situations, such as disturbances, distractions, and pathologies, that interfere with execution of an intended movement. One can assume that a person who is familiar with a particular task will try to do the same thing each time, provided all other conditions are equal. This presumes that what people do is on average what they intended to do and that their goal remains unchanged. However, if there is a disturbance, it is unclear what happens to a person's intentions because action is decoupled from intention. Challenges arise in human-machine interactions because intention and changes in intention cannot be easily discerned. Hence a comprehensive method for understanding what someone means to do (rather than what they actually do or claim to have wanted to do) as they interact within uncertain environments will prove useful in any area where a person might be thwarted from their intent such as national security, operation of machinery, athletic and music performance, rehabilitation, human augmentation, and artificial intelligence.

It is important to distinguish our use of the term intent from other groups that consider \textit{motivation}\cite{mcclelland1985motives, rawolle2013relationships}, \textit{cost}\cite{todorov2002optimal, flash1985coordination}, or \textit{goal selection decisions}\cite{ziebart2010modeling}. In typical motor control studies, subjects are motivated to complete the experiment in a timely fashion and are usually explicitly provided targets to be reached. Classification of intent is prevalent in both lower \cite{strausser2011development, hargrove2013robotic} and upper limb\cite{englehart2003robust, young2012improving} prosthetics where hybrid control algorithms select from among a set of actions (walking/standing/flexion/extension/etc). While subjects may be \textit{motivated} to complete experiments with minimal effort/cost and their \textit{goal} may be to reach a target, here we use intent to describe the course of action (i.e., the trajectory of the arm) taken in service of goals and motives and not the goals nor the motives themselves. Particularly of interest is the intended course of action (i.e, the intended trajectory), even when the actual movement is disturbed and hence no longer matches the intent.

Attempts to deduce motor intent in the past have focused on the assumed spring-like properties of human muscles. Springs produce a force according to their impedance and stretch. By measuring force, impedance, and position, Gomi and Kawato\cite{gomi1997human} were able to deduce stretch and thereby infer the muscle's equilibrium point. Supporters of the ``$\lambda$ model''\cite{feldman1995origin} hypothesized that this equilibrium point represented the intent of a movement. Upon Gomi and Kawato's inspection of the equilibrium point as it evolved in time, it was clear that it was highly complex and often not anatomically realizable. Therefore, it could not well-represent the intent of a simple reaching movement. We propose a revised and more general method, \textit{intent extraction}, which calculates the intended trajectory even in the face of disturbances by relating environmental interaction and state to intent. This method affords new ways to 1) test and falsify candidate models of motor control, 2) study the timing and composition of alterations to intent during movement, and 3) understand how intent might be altered by pathology after neural insult. This method describes a class of filters that map a disturbance, process, actuator, and realized trajectory to an intended trajectory. In human reaching, these initial ingredients are available: the process is a dynamic model of the arm, the actuators (muscles) are assumed to act as viscoelastic torque generators, the human interface forces are measured by a sensor, and the realized trajectory is measured.

Here we develop and analyze the mathematics underlying intent extraction, a method to determine one's intent. We focused, for simplicity in this initial study, on planar targeted arm reaching actions. First, we produce synthetic reaching data using a forward model so that we can examine the quality and uncertainty of intent extraction. We then explore sensitivity and failure modes, demonstrating the accuracy and utility of this method. We show efficacy on a preliminary dataset of healthy reaching movements disrupted by occasional, unpredictable disturbances. Finally, we demonstrate the viability of this method for providing a better understanding of what a person means to do when they move and are disturbed, showing how one's intent may change following a disturbance.

@@

For typical dynamic simulations, in order to determine the trajectory of the system (i.e., the forward dynamics problem), this second-order differential equation is solved by numerical integration to determine the solution to the initial value problem in time. This first involves algebraic manipulation to solve for $\ddot{q}$, and then integration to determine the state trajectory. Intent determination takes the novel approach of instead solving for $\ddot{q}_d$:
\begin{equation}
\ddot{q}_d=\hat{M}(q_d)^{-1}\left\{M(q)\ddot{q}+G(q,\dot{q})+E-[\hat{G}(q_d,\dot{q}_d)+\hat{E}+K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})]\right\}
\end{equation}  
$\ddot{q}_d$ can then be integrated to determine the intended state trajectory $q_d(t)$. This relies on many assumptions: The model of plant and controller must be accurate and precise. The initial conditions must be available and accurate. The mass matrix estimate, $\hat{M}$, must be invertible. Externally-applied force must be precisely and accurately measured. If all of these conditions are met, then the system yields an accurate estimate of the intent.

@@

\subsection*{Generation of synthetic data using simple dynamic simulation}
While the derivation above is general, testing its application to human reaching requires choices for plant and actuator models. Here we show the results from our choice of Burdet and colleagues \cite{burdet2006stability}. Appropriate plant modeling is well-understood: measured arm segment lengths and self-reported body mass are converted into the plant's inertial, centripetal, and Coriolis terms. Anatomical landmarks and values from Dempster \cite{dempster1955space} and Winter \cite{winter2009biomechanics} relate body mass to limb mass, limb length to limb center of mass, and limb mass and length to moment of inertia. Desired trajectory in time was simulated as a minimum jerk, 5th order polynomial of duration 700 milliseconds starting and ending with zero velocity and acceleration \cite{flash1985coordination}. For the actuator, we choose the model of Burdet et al. \cite{burdet2006stability}. A scalar multiple of their modeled feedback torque, determined by linear regression, was used to account for any task-dependent differences in impedance \cite{franklin2003adaptation} while leaving the shape and structure of the impedance unchanged. In this  model, musculoskeletal stiffness was linearly related to torque. Like Burdet et al. \cite{burdet2006stability}, we approximated muscle torque by looking backward in time 500 microseconds. Note that this backward look led to negligible (sub-micrometer mean unsigned error) discrepancies in subsequent intent estimates.

@@

\subsection*{Indices of variance-based sensitivity}
To ensure a properly spaced and efficient evaluation of sensivity, we employed Sobol-distributed matrices, generated using MATLAB's \textit{sobolset()} function and converted to parameter distributions (see Table~\ref{tab:parameters}) using inverse cumulative probability density. Distributions were clamped to the range of $\pm3$ standard deviations in order to limit parameters to a realistic range. Combinations and calculations were made as prescribed by Saltelli et al. \cite{saltelli2010variance} to arrive at direct and total sensitivity indices.

@@

\subsection*{Apparatus}
A planar manipulandum (described in Patton and Mussa-Ivaldi \cite{patton2004robot}) was programmed to compensate and minimize any friction or mass. The MATLAB XPC-TARGET package \cite{MATLAB:2008} was used to render this force environment at 1000 Hz and data were collected at 1000 Hz.  Visual feedback was performed at 60 Hz using OpenGL. Closed-loop data transmission time (position measurement to completed rendering to recognition of rendering by the position measurement system) was less than 8 milliseconds, ensuring a visual delay less than one 60 Hz frame. Because force sensors tend to drift, we performed a linear re-zeroing procedure between each motion to assure unbiased measurements.

@@

\subsection*{Statistical Analysis}
The MATLAB statistics toolbox package \cite{MATLAB:2008} was used to perform comparisons among subjects and trajectory sources using ANOVA with Tukey-Kramer post-\textit{hoc} test for group differences. We used an alpha level of 0.05 for declaring statistical significance.

@@

\subsection*{Parameter Sensitivity}
Variance in intended trajectories due to estimated uncertainty in model parameters was lower than the natural variation in undisturbed motions. Simulated point-to-point reaches disturbed by either filtered white noise forces or a pulse force perpendicular to the direction of the reach (Figure 1) were extracted in the presence of 220,000 variations upon the parameters (Table 1) according to the methods of Saltelli et al. \cite{saltelli2010variance}. Expected variance due to parameter uncertainty was $0.86  \mu m^2$ for pulse forces and $1.5 \mu m^2$ for filtered white forces. Variance in recorded undisturbed point-to-point reaching under the same time and reach distance conditions was $2.93 \mu m^2$. These variances all fall beneath the instantaneous accuracy of the robot recording the motion (tenths of a millimeter).

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Our actions often do not match our intentions when there are external disturbances. We derive a novel modeling approach for determining motor intent from a reaching motion that is disturbed by an unexpected force. We first show how to mathematically invert both feedforward (predictive) and feedback controls to obtain intent. We next examine the model’s sensitivity to parameter uncertainty, and test the approach using human reaching data. We found that sensitivity to parameter inaccuracies was below one square millimeter, and hence much lower than the inaccuracies seen in actual human motions. Further, the largest sensitivity arose mainly from uncertainty in our assumptions of joint stiffness. We also tested the hypothesis that this would predict straight-line intent in human targeted reaching movements, even when disturbed by various pulses of force. As expected, the intended trajectory showed no change from undisturbed reaching for 300 ms after the disturbance onset. Beyond this point, however, we detected a change in intent, even when the intention was near the target.  Knowing such an intent signal is broadly applicable in areas such as in enhanced human-machine interaction, the study of impaired intent in neural disorders, the real-time determination (and manipulation) of error in training, and in any complex system that embodies planning, such as brain machine interfaces, team sports, crowds, or swarms. In addition, observing intent as it changes might act as a window into the mechanisms of planning, correction, and learning.

@@

\section*{Introduction}
There are many situations, such as disturbances, distractions, and pathologies, that interfere with execution of an intended movement. One can assume that a person who is familiar with a particular task will try to do the same thing each time, provided all other conditions are equal. This presumes that what people do is on average what they intended to do and that their goal remains unchanged. However, if there is a disturbance, it is unclear what happens to a person's intentions because action is decoupled from intention. Challenges arise in human-machine interactions because intention and changes in intention cannot be easily discerned. Hence a comprehensive method for understanding what someone means to do (rather than what they actually do or claim to have wanted to do) as they interact within uncertain environments will prove useful in any area where a person might be thwarted from their intent such as crowd control, operation of machinery, athletic and music performance, rehabilitation, human augmentation, and artificial intelligence.

It is important to distinguish intended trajectory from \textit{motivation}\cite{mcclelland1985motives, rawolle2013relationships}, \textit{cost}\cite{todorov2002optimal, flash1985coordination}, or \textit{goal selection decisions}\cite{ziebart2010modeling}. In typical motor control studies, subjects are motivated to complete the experiment in a timely fashion and are usually explicitly provided targets to be reached. Classification of intent is prevalent in both lower \cite{strausser2011development, hargrove2013robotic} and upper limb\cite{englehart2003robust, young2012improving} prosthetics where hybrid control algorithms select from among a set of actions (walking/standing/flexion/extension/etc). While subjects may be \textit{motivated} to complete experiments with minimal effort/cost and their \textit{goal} may be to reach a target, here we use intent to describe the course of action (i.e., the trajectory of the arm) taken in service of goals and motives and not the goals nor the motives themselves. Particularly of interest is the intended course of action (i.e, the intended trajectory), even when the actual movement is disturbed and hence no longer matches the intent.

Attempts to deduce motor intent in the past have focused on the assumed spring-like properties of human muscles. Springs produce a force according to their impedance and stretch. By measuring force, impedance, and position, Gomi and Kawato\cite{gomi1997human} were able to deduce stretch and thereby infer the muscle's equilibrium point. Supporters of the ``$\lambda$ model''\cite{feldman1995origin} hypothesized that this equilibrium point represented the intent of a movement even though it did not consider mass as a feedforward controller would. Upon Gomi and Kawato's inspection of the equilibrium point as it evolved in time, it was clear that it was highly complex and often not anatomically realizable. Therefore, it could not well-represent the intent of a simple reaching movement. We propose a revised and more general method, \textit{intent extraction}, which calculates the intended trajectory even in the face of disturbances by relating environmental interaction and state to intent. This method affords new ways to 1) test and falsify candidate models of motor control, 2) study the timing and composition of alterations to intent during movement, and 3) understand how intent might be altered by pathology after neural insult. This method describes a class of filters that map a disturbance, process, actuator, and realized trajectory to an intended trajectory. In human reaching, these initial ingredients are available: the process is a dynamic model of the arm, the actuators (muscles) are assumed to act as viscoelastic torque generators, the human interface forces are measured by a sensor, and the realized trajectory is measured.

Here we develop and analyze the mathematics underlying intent extraction, a method to determine one's intended trajectory. We focused, for simplicity in this initial study, on planar targeted arm reaching actions. First, we produce synthetic reaching data using a forward model so that we can examine the quality and uncertainty of intent extraction. We then explore sensitivity and failure modes, demonstrating the accuracy and utility of this method. We show efficacy on a preliminary dataset of healthy reaching movements disrupted by occasional, unpredictable disturbances. Finally, we demonstrate the viability of this method for providing a better understanding of what a person means to do when they move and are disturbed, showing how one's intended trajectory may change following a disturbance.

@@

For typical dynamic simulations, in order to determine the trajectory of the system (i.e., the forward dynamics problem), this second-order differential equation is solved by numerical integration to determine the solution to the initial value problem in time. This first involves algebraic manipulation to solve for $\ddot{q}$, and then integration to determine the state trajectory. Intent determination takes the novel approach of instead solving for $\ddot{q}_d$:
\begin{equation}
\ddot{q}_d=\hat{M}(q_d)^{-1}\left\{M(q)\ddot{q}+G(q,\dot{q})+E-[\hat{G}(q_d,\dot{q}_d)+\hat{E}+K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})]\right\}
\end{equation}  
$\ddot{q}_d$ can then be integrated numerically using a differential equation solver to determine the intended state trajectory $q_d(t)$. This relies on many assumptions: The model of plant and controller must be accurate and precise. The initial conditions must be available and accurate. The mass matrix estimate, $\hat{M}$, must be invertible. Externally-applied force must be precisely and accurately measured. If all of these conditions are met, then the system yields an accurate estimate of the intent.

@@

\subsection*{Dynamic simulation of arm and intended trajectories}
While the derivation above is general, testing its application to human reaching requires choosing plant and actuator models, which themselves require physical parameters. Appropriate plant modeling is well-understood: measured arm segment lengths and self-reported body mass are converted into the plant's inertial, centripetal, and Coriolis terms. Anatomical landmarks and values from Dempster \cite{dempster1955space} and Winter \cite{winter2009biomechanics} relate body mass to limb mass, limb length to limb center of mass, and limb mass and length to moment of inertia. For the actuator, we choose the model and parameters of Burdet et al. \cite{burdet2006stability}. A scalar multiple of their modeled feedback torque, determined by linear regression, was used to account for any task-dependent differences in impedance \cite{franklin2003adaptation} while leaving the shape and structure of the impedance unchanged. Desired trajectory in time was simulated as a minimum jerk, 5th order polynomial of duration 700 milliseconds starting and ending with zero velocity and acceleration \cite{flash1985coordination}.  In this  model, musculoskeletal stiffness was linearly related to torque. Like Burdet et al. \cite{burdet2006stability}, we approximated muscle torque by looking backward in time 500 microseconds. Note that this backward look led to negligible (sub-micrometer mean unsigned error) discrepancies in subsequent intent estimates.

@@

\subsection*{Indices of variance-based sensitivity}
To ensure a properly spaced and efficient evaluation of sensivity, we employed Sobol-distributed matrices, generated using MATLAB's \textit{sobolset()} function and converted to parameter distributions (see Table~\ref{tab:parameters}) using inverse cumulative probability density. Nominal parameters values were mostly taken from taken Dempster \cite{dempster1955space} and Burdet et al. \cite{burdet2006stability} with typical values where subject-specific parameters (height and weight) and direct where necessary (force sensor noise and drift). Standard deviations reflect expected variation in repeated measurement (ie. height and weight) or measured variance (ie. trial-to-trial variation in sensor noise or uncertainty in mass ratios reported by Dempster \cite{dempster1955space}). Distributions were clamped to the range of $\pm3$ standard deviations in order to limit parameters to a realistic range. Combinations and calculations were made as prescribed by Saltelli et al. \cite{saltelli2010variance} to arrive at direct and total sensitivity indices.

@@

\subsection*{Apparatus}
A planar manipulandum (described in Patton and Mussa-Ivaldi \cite{patton2004robot}) was programmed to compensate and minimize any friction or mass. The MATLAB XPC-TARGET package \cite{MATLAB:2008} was used to render this force environment at 1000 Hz and data were collected at 1000 Hz.  Visual feedback of hand position was performed at 60 Hz using OpenGL. Closed-loop data transmission time (position measurement to completed rendering to recognition of rendering by the position measurement system) was less than 8 milliseconds, ensuring a visual delay less than one 60 Hz frame. Because force sensors tend to drift, we performed a linear re-zeroing procedure between each motion to assure unbiased measurements.

@@

\subsection*{Statistical Analysis}
The MATLAB statistics toolbox package \cite{MATLAB:2008} was used to perform comparisons of maximum deviation among subjects and trajectory sources (hand and deviation) using ANOVA with Tukey-Kramer post-\textit{hoc} test for group differences. We used an alpha level of 0.05 for declaring statistical significance.

@@

\subsection*{Parameter Sensitivity}
Variance in intended trajectories due to estimated uncertainty in model parameters was lower than the natural variation in undisturbed motions. Simulated point-to-point reaches disturbed by either filtered white noise forces or a pulse force perpendicular to the direction of the reach (Figure 1) were extracted in the presence of 220,000 variations upon the parameters (Table 1) according to the methods of Saltelli et al. \cite{saltelli2010variance}. Expected variance due to parameter uncertainty was $0.86  mm^2$ for pulse forces and $1.5 mm^2$ for filtered white forces. Variance in recorded undisturbed point-to-point reaching under the same time and reach distance conditions was $2.93 mm^2$. These variances describe mean unsigned error, but do not address maximum error.
